---
title: "Other Graphs"
author: "David Asare Kumi"
date: "4/3/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


## Correlation Matrix

- Here we are not going to focus on data that is categorical.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
data(mtcars)
head(mtcars)

mt_cor<-cor(mtcars)
round(mt_cor,2)

```

## Graph the correlation matrix

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(corrplot)
data(mtcars)

corrplot(mt_cor)

```

## Some options of the corrplot() function

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(corrplot)
data(mtcars)

corrplot(mt_cor,method="shade",shade.col = NA,tl.col = "black",tl.srt = 45)

```

## Display tables representing the correlation

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(corrplot)
data(mtcars)

# Generate a lighter palette
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))

corrplot(mt_cor, method="shade", shade.col=NA, tl.col="black", tl.srt=45,
col=col(200), addCoef.col="black", addcolorlabel="no", order="AOE")

```

## Display tables representing the correlation(alternative)

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(corrplot)
data(mtcars)

# Generate a lighter palette
colors <- c("lightblue","orange","pink","yellow","yellowgreen","white","red","maroon","fuchsia",
            "aqua")

corrplot(mt_cor, method="shade", shade.col=NA, tl.col="black", tl.srt=45,
col=colors(200), addCoef.col="white", addcolorlabel="no", order="AOE")

```

## Display tables representing the correlation(alternative with method=ellipse)

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(corrplot)
data(mtcars)

# Generate a lighter palette
colors <- c("lightblue","orange","pink","yellow","yellowgreen","white","red","maroon","fuchsia",
            "aqua")

corrplot(mt_cor, method="ellipse", shade.col=NA, tl.col="black", tl.srt=45,
col=colors(200), addCoef.col="white", addcolorlabel="no", order="AOE")

```

## Display tables representing the correlation(alternative with order=hclust)

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(corrplot)
data(mtcars)

# Generate a lighter palette
colors <- c("lightblue","orange","pink","yellow","yellowgreen","white","red","maroon","fuchsia",
            "aqua")

corrplot(mt_cor, method="shade", shade.col=NA, tl.col="black", tl.srt=45,
col=colors(200), addCoef.col="white", addcolorlabel="no", order="hclust")

```

## Options for corrplot()

- type={"lower" | "upper"} Only use the lower or upper triangle.

- diag=FALSE Don’t show values on the diagonal.

- addshade="all" Add lines indicating the direction of the correlation.

- shade.col=NA Hide correlation direction lines.

- method="shade" Use colored squares.

- method="ellipse" Use ellipses.

- addCoef.col="color" Add correlation coefficients, in color.

- tl.srt="number" Specify the rotation angle for top labels.

- tl.col="color" Specify the label color.

- order={"AOE" | "FPC" | "hclust"} Sort labels using angular order of eigenvectors, first principle component, or hierarchical clustering

## Creating a Network Graph

- Use the igraph package. To create a graph, pass a vector containing pairs of items to
graph(), then plot the resulting object

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}

library(igraph)
library(ggplot2)
library(plotly)

# Specify edges for a directed graph
gd <- graph(c(1,2, 2,3, 2,4, 1,4, 5,5, 3,6))
plot(gd)

# For an undirected graph
gu <- graph(c(1,2, 2,3, 2,4, 1,4, 5,5, 3,6), directed=FALSE)
# No labels
plot(gu, vertex.label=NA)

```


