---
title: "Bar Graph2"
author: "David Asare Kumi"
date: "4/2/2020"
output: html_document
---


## Basic Bar Graphs

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(pg_mean)

g<-ggplot(pg_mean,aes(x=group,y=weight))+geom_bar(stat="identity",fill="yellow",color="black")+
  ggtitle("Barplot of Group and Weight")
ggplotly(g)

```

## Grouping Bars Together

- Group bars together by a second variable.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(cabbage_exp)

g<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",position="dodge") + xlab("Date")+ ylab("Weight")+ggtitle("Group Barplot of Date and Weight")
ggplotly(g)

```

- It is important to note that, if you don't use position="dodge", you will end up with a stacked bar graph.

- Variables mapped to the fill color must be categorical or factor variables and not continuous variables.

## Set the colors

1. Use scale_fill_brewer()

2. Use scale_fill_manual()

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(cabbage_exp)

g<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",position="dodge",
color="black")+scale_fill_brewer(palette="Dark2") + xlab("Date")+ ylab("Weight")+ggtitle("Group Barplot of Date and Weight")
ggplotly(g)

```

- If there are missing observations in the categorical variable, the bar will be missing and the neighbouring bars will expand to fill that space.

- If your data has this issue, you can manually make an entry for the missing factor level combination with an NA for the y variable.

## Making a bar graph of counts

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(ggplot2)
library(plotly)
data(diamonds)

g<-ggplot(diamonds,aes(x=cut,fill=cut))+ geom_bar()+xlab("Cut")+ylab("Count")+
  ggtitle("Barplot showing count/frequency of cut ")
ggplotly(g)

```

- In this example, with geom_bar(), the default behaviour is to use stat="bin" which counts up the number of cases for each group.

- If we use a continuous variable on the x-axis, we get a histogram.

## Continuous variable resulting in a histogram instead of a bar graph

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(ggplot2)
library(plotly)
data(diamonds)

g<-ggplot(diamonds,aes(x=carat))+ geom_bar(stat="bin")+xlab("Carat")+ylab("Count")+
  ggtitle("Barplot showing count/frequency of Carat ")
ggplotly(g)

```

## Using colors in a bar graph


```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(uspopchange)

upc<-subset(uspopchange,rank(Change)>40)

g<-ggplot(upc,aes(x=Abb,y=Change,fill=Region))+geom_bar(stat="identity")
ggplotly(g)

```

## Set colors manually

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(uspopchange)

upc<-subset(uspopchange,rank(Change)>40)

g<-ggplot(upc,aes(x=reorder(Abb,Change),y=Change,fill=Region))+geom_bar(stat="identity",
color="black")+scale_fill_manual(values=c("blue","yellow"))+xlab("State")
ggplotly(g)

# you can set the color using scale_fill_manual(values=c("#669933","#FFCC66"))

```

## Coloring Negative and Positive bars differently

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(climate)

csub<-subset(climate,Source=="Berkeley" & Year>=1900)
csub$pos<-csub$Anomaly10y>=0

g<-ggplot(csub,aes(x=Year,y=Anomaly10y,fill=pos))+geom_bar(stat="identity",position="identity")+
  scale_fill_manual(values=c("yellow","blue"))+ggtitle("Barplot of Year and Anomaly10y")
ggplotly(g)

```

- We can change the colors with scale_fill_manual() or scale_fill_brewer()

- Remove the legend with guide=FALSE - this is deprecated.

- During the plot, you can turn off legends by using the argument show.legend=FALSE in the geom_bar().

- After the plot creation, it’s possible to remove the legend using the code p + theme(legend.position="none").

- Add a thin black line around each of the bars by setting color="black" and specify size which is the thickness of the outline in millimetres.


```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(climate)

csub<-subset(climate,Source=="Berkeley" & Year>=1900)
csub$pos<-csub$Anomaly10y>=0

g<-ggplot(csub,aes(x=Year,y=Anomaly10y,fill=pos))+geom_bar(stat="identity",position="identity",
color="black",size=0.25)+scale_fill_manual(values=c("#CCEEFF","#FFDDDD"))+ggtitle("Barplot of Year and Anomaly10y")+theme(legend.position="none")
ggplotly(g)

```

## Adjusting Bar Width and Spacing

1. Set width in geom_bar(). The default value is 0.9. Larger values make the bars wider and smaller values make the bars narrower.

2. The maximum width is 1.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#Standard width
library(gcookbook)
library(ggplot2)
library(plotly)
library(gridExtra)
data(pg_mean)

g1<-ggplot(pg_mean,aes(x=group,y=weight,fill=group))+geom_bar(stat="identity")+xlab("Group")+
  ylab("Weight")+ggtitle("Barplot of Group and Weight")
ggplotly(g1)

#For narrow bars
g2<-ggplot(pg_mean,aes(x=group,y=weight,fill=group))+geom_bar(stat="identity",width=0.5)+xlab("Group")+ ylab("Weight")+ggtitle("Barplot of Group and Weight")
ggplotly(g2)

#For wider bars
g3<-ggplot(pg_mean,aes(x=group,y=weight,fill=group))+geom_bar(stat="identity",width=1)+xlab("Group")+ ylab("Weight")+ggtitle("Barplot of Group and Weight")
ggplotly(g3)

grid.arrange(g1,g2,g3,ncol=3)

```


## Create Space in grouped bars and some space between the bars

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(gridExtra)
data(cabbage_exp)

#space between grouped bars
g1<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",width=0.5,
position="dodge")+ggtitle("Barplot of Date and Weight")
ggplotly(g1)

#some space between the bars
g2<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",width=0.5,
position=position_dodge(0.7))+ggtitle("Barplot of Date and Weight")
ggplotly(g2)

#apply gridExtra package
grid.arrange(g1,g2,ncol=2)

```

- position="dodge" has a default value of 0.9. If we want to set a specific value then we use position=position_dodge() for the bars.

- The default width value for position _dodge() is the same as width in geom_bar().

## Making a stacked bar graph plus the label in the middle of the stack

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(cabbage_exp)

g<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity")+xlab("Date")+
  ylab("Weight")+ggtitle("Stacked bar graph of Date and Weight")+geom_text(aes(label=Weight),
                                      size=3,position=position_stack(vjust=0.5))
ggplotly(g)

```

- Note that x is a categorical or factor variable and fill is also mapped onto another categorical or factor variable.

## Reverse the order of the legend

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
data(cabbage_exp)

g<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity")+
guides(fill=guide_legend(reverse=TRUE))+ xlab("Date")+ylab("Weight")+ggtitle("Stacked bar graph of Date and Weight")
ggplotly(g)

```

- The structure of the cabbage_exp data.

1. There are three levels of Date.

2. There are two levels of Cultivar.

## To reverse the stacking order

- Specify order = desc() in the aesthetic mapping.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
data(cabbage_exp)

g<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar,order=desc(Cultivar)))+
  geom_bar(stat="identity")+xlab("Date")+ylab("Weight")+ggtitle("Stacked bar graph of Date and Weight")
ggplotly(g)

```

## For a more polished graph

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
data(cabbage_exp)

g<-ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",color="black")+ guides(fill=guide_legend(reverse=TRUE))+scale_fill_brewer(palette="Pastel1")+xlab("Date")+ylab("Weight")+ggtitle("Stacked bar graph of Date and Weight")
ggplotly(g)

```

## Making a proportional stacked bar graph (%age)

- Scale the data to 100% within each stack.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
data(cabbage_exp)

#Do a group wise transform splitting on "Date".

ce<-ddply(cabbage_exp,"Date",transform,percent_weight=Weight/sum(Weight)*100)

g<-ggplot(ce,aes(x=Date,y=percent_weight,fill=Cultivar))+geom_bar(stat="identity")
ggplotly(g)

```

## Adding labels to a bar graph

- Add geom_text to your graph and set vjust i.e. vertical justification which enables you to move the text above or below the tops of the bars.


```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
library(gridExtra)
data(cabbage_exp)

#below the top
g1<-ggplot(cabbage_exp,aes(x=interaction(Date,Cultivar),y=Weight))+geom_bar(stat="identity")+
  geom_text(aes(label=Weight),vjust=1.5,color="white")

#above the top
g2<-ggplot(cabbage_exp,aes(x=interaction(Date,Cultivar),y=Weight))+geom_bar(stat="identity")+
  geom_text(aes(label=Weight),vjust=-0.2) 

grid.arrange(g1,g2,ncol=2)


```

- Consider your ylimt when you want to put the text on top of the bars. Y upper limit must be a bit higher.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
library(gridExtra)
data(cabbage_exp)

#Adjjust y limits to be a little higher
g1<-ggplot(cabbage_exp,aes(x=interaction(Date,Cultivar),y=Weight))+geom_bar(stat="identity")+
geom_text(aes(label=Weight),vjust=-0.2)+ylim(0,max(cabbage_exp$Weight)*1.05)

#map y position slightly above the top of the bar
g2<-ggplot(cabbage_exp,aes(x=interaction(Date,Cultivar),y=Weight))+geom_bar(stat="identity")+
geom_text(aes(y=Weight+0.1,label=Weight))

grid.arrange(g1,g2,ncol=2)

```

## For grouped bar graph

- Use position =position_dodge() and give it a value for the dodging width- default is 0.9.

- Specify size - default is 5.

- make it smaller by using 3.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
library(gridExtra)
data(cabbage_exp)

ggplot(cabbage_exp,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",position="dodge")+
  geom_text(aes(label=Weight),vjust=1.5,color="white",position=position_dodge(0.9),size=4)

```

## Putting labels on stacked bars

- Find the cumulative sum for each stack.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
library(gridExtra)
data(cabbage_exp)

#sort the data
ce<-arrange(cabbage_exp,Date,Cultivar)
#get the cumulative sum
ce<-ddply(ce,"Date",transform,label_y=cumsum(Weight))
ce

g1<-ggplot(ce,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity")+geom_text(aes(y=label_y,
    label=Weight),vjust=1.5,color="white")

#placing the label in the middle of each stacked bar
ce<-arrange(cabbage_exp,Date,Cultivar)

#calculate y position placing it in the middle
ce<-ddply(ce,"Date",transform,label_y=cumsum(Weight)-0.5*Weight)

g2<-ggplot(ce,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity")+
  geom_text(aes(y=label_y,label=Weight),color="white")

grid.arrange(g1,g2,ncol=2)

```

## For a more polished graph

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(gcookbook)
library(ggplot2)
library(plotly)
library(plyr) # needed for desc()
library(gridExtra)
data(cabbage_exp)

#placing the label in the middle of each stacked bar
ce<-arrange(cabbage_exp,Date,Cultivar)

#calculate y position placing it in the middle
ce<-ddply(ce,"Date",transform,label_y=cumsum(Weight)-0.5*Weight)


ggplot(ce,aes(x=Date,y=Weight,fill=Cultivar))+geom_bar(stat="identity",color="black")+
  geom_text(aes(y=label_y,label=paste(format(Weight,nsmall=2),"kg")),size=4)+
  guides(fill=guide_legend(reverse=TRUE))+scale_fill_brewer(palette="Pastel1")

```

## Add data label to simple bargraph

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(datasets)
library(ggplot2)
library(plotly)
data(diamonds)


g<-ggplot(diamonds,aes(x=cut,fill=cut))+ geom_bar()+xlab("Cut")+ylab("Count")+
  ggtitle("Barplot showing count/frequency of cut ")+geom_text(stat="count",aes(label=..count..),
  vjust=-0.2)
ggplotly(g)

```